---
title: "VCF mtDNA check "
author: "Hansi"
date: "July 28, 2025"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
---

```{r setup, include=FALSE}
# Global options for all chunks
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

library(vcfR)
library(seqinr)
library(dplyr)
library(stringr)
library(ggplot2)
library(networkD3)


source("../../scripts/haplogroup_processing.R")

```


# Introduction
This R Markdown document performs a series of tasks on a VCF file, a FASTA reference, and a Haplogrep input file. The process involves: 
    - Checking the VCFs reference alleles against a FASTA file.
    - Swapping variants where the ALT allele matches the reference.
    - Generating a list of variant positions to update the Range column in a Haplogrep input file.
    - Comparing the haplogroup performance to some common genotyping arrays (best-case-scenarios)




# Parameters 

```{r }


# Replace with your VCF file path
#
# Other 1000G microarrays (Affy 6, axiom, or Illumina Omni to be found here  https://www.internationalgenome.org/category/omni/ 


vcf_file <- "data/example-microarray.vcf"

fasta_file <- "../../reference/mtDNA/chrM.fasta" 

###HaploGrep Phylotree 17 files
#------------------------------
#phylotreeVersion<-"17"
#haplogrep_input_file <- "../../reference/mtDNA/phylotree17.hsd" 
#weights<-"../../reference/mtDNA/phylotree17_weights.txt"

###HaploGrep Phylotree 17 FU1 files 
#----------------------------------
phylotreeVersion<-"17_FU1"
haplogrep_input_file <- "../../reference/mtDNA/phylotree17FU1_mod.hsd" 
weights<-"../../reference/mtDNA/phylotree17FU1_weights.txt"

#small test data not to run over the entire phylogenetic tree
#haplogrep_input_file <- "../../reference/mtDNA/phylotree17FU1_weights.txt" 

output_haplogrep_file <- "output/modified_haplogrep_input.txt" 
haplogroup_distcheck_file <-"output/haplogroup.distcheck.txt" 
haplogroup_distcheck_result_file <-"output/haplogroup.distcheckResult.txt"

#for simple comparison reasons - check with Illumina GSA (best) and Illumina Omni (low-performer)

array_omni_file <- "data/arrays/InfiniumOmni2-5-8v1-5_A1.txt"
array_gsa_file <-  "data/arrays/GSA-24v3-0_A1.txt"
array_mega_file <- "data/arrays/Multi-EthnicGlobal_D.grch18.chrM.txt"
array_axiom_ukb_file <- "data/arrays/Axiom_UKB.txt"
#array_gsa4_file <-  "data/arrays/GSA-PGx-48v4.MT.txt"
#array_mega_A1_file <-"data/arrays/Multi-EthnicGlobal_A1-b37-strand_MT.txt"
#array_mega_D1_file <-"data/arrays/Multi-EthnicGlobal_D1-b37-strand_MT.txt"

```


# Check reference 

```{r, warning=FALSE, echo=FALSE }

# ==============================================================================
# Step 1: Check Reference Variants against FASTA and Plot
# ==============================================================================
message("Step 1: Checking reference variants against FASTA...")

# --- Step 0: Load VCF and FASTA ---
vcf <- read.vcfR(vcf_file, verbose = FALSE)
weightsPhylo<-read.delim(weights)

# Robust FASTA file reader to handle line breaks
fasta_lines <- readLines(fasta_file)
fasta_header_idx <- grep("^>", fasta_lines)
fasta_sequence_lines <- fasta_lines[-fasta_header_idx]
fasta_sequence <- paste(fasta_sequence_lines, collapse = "")
fasta_sequence <- toupper(str_replace_all(fasta_sequence, "[^A-Z]", ""))

# --- Helper function to get a base from the fasta object ---
# This version works with a single sequence string
get_fasta_base <- function(fasta_sequence, chrom, pos) {
  
  if (nchar(fasta_sequence) == 0) {
    stop("FASTA sequence is empty. Please check your FASTA file for sequence data.")
  }

  if (pos > nchar(fasta_sequence) || pos < 1) {
    warning(paste("Position", pos, "is out of bounds for chromosome", chrom, ". Skipping variant."))
    return(NA)
  }
  
  return(substr(fasta_sequence, pos, pos))
}

# Get reference alleles from FASTA based on VCF coordinates
vcf_fix <- vcf@fix
if (nrow(vcf_fix) == 0) {
    stop("VCF file appears to be empty or contains no variants. Cannot proceed.")
}

fasta_ref_alleles <- mapply(get_fasta_base,
                            chrom = vcf_fix[, "CHROM"],
                            pos = as.numeric(vcf_fix[, "POS"]),
                            MoreArgs = list(fasta_sequence = fasta_sequence))

# Filter out variants where no FASTA base could be found
valid_variants <- !is.na(fasta_ref_alleles)
if (sum(valid_variants) == 0) {
    stop("! No valid variants remain after matching to the FASTA reference.")
}
vcf_fix <- vcf_fix[valid_variants, ]
fasta_ref_alleles <- fasta_ref_alleles[valid_variants]
vcf <- vcf[valid_variants, ]

# Create a data frame for comparison
ref_comparison_df <- data.frame(
    POS= as.integer(vcf_fix[, "POS"]),
    VCF_ID = as.character(vcf_fix[, "ID"]),
    VCF_REF = as.character(vcf_fix[, "REF"]),
    VCF_ALT = as.character(vcf_fix[, "ALT"]),
    FASTA_REF = as.character(fasta_ref_alleles),
    stringsAsFactors = FALSE
)



# Count occurrences for plotting
plot_data <- ref_comparison_df %>%
    dplyr::mutate(Match = ifelse(VCF_REF == FASTA_REF, "Matches FASTA", "Mismatches FASTA")) %>%
    dplyr::group_by(VCF_REF, Match) %>%
    dplyr::summarise(Count = n(), .groups = 'drop')

ggplot_ref_check <- ggplot(plot_data, aes(x = VCF_REF, y = Count, fill = Match)) +
    geom_bar(stat = "identity", position = "dodge") +
    geom_text(aes(label = Count), vjust = -0.5, position = position_dodge(width = 0.9)) +
    labs(
        title = "Reference Allele Consistency with FASTA",
        x = "VCF Reference Allele",
        y = "Number of Occurrences",
        fill = "Comparison"
    ) +
    theme_minimal() +
    scale_fill_manual(values = c("Matches FASTA" = "lightgreen", "Mismatches FASTA" = "darkred")) + theme(legend.position="bottom")

print(ggplot_ref_check)
message("Step 1: Reference check complete and plot generated.")



```

# Check Phylogenetic relevance of POS

```{r, warning=FALSE, echo=FALSE }



ref_comparison_df$VARIANT<-paste(ref_comparison_df$POS, ref_comparison_df$VCF_ALT, sep="")
checkPhyloWeights<-merge(ref_comparison_df, weightsPhylo, by.x="VARIANT", by.y="Mutation", all.x=TRUE)

checkPhyloWeights$countN[is.na(checkPhyloWeights$countN)] <- 0

ggplot(checkPhyloWeights, aes(x=POS, y=countN)) + geom_point(aes(color=countN>0)) + facet_wrap(~countN>0, ncol=1) + labs(title="Phylogenetically relevant positions along chrM", subtitle="TRUE=in Phylotree, FALSE=not in Phylotree") + theme_minimal() + scale_color_manual(values = c("TRUE" = "lightgreen", "FALSE" = "darkred")) + theme(legend.position="bottom")

ggplot(checkPhyloWeights, aes(x=countN, fill = as.factor(countN == 0))) + geom_bar() +
    theme_minimal() +
    scale_fill_manual(
        name = "Positions phylogenetically relevant",
        values = c("TRUE" = "darkred", "FALSE" = "lightgreen"),
        labels = c("TRUE" = "unknown", "FALSE" = "relevant for haplogroup")
    ) + theme(legend.position="bottom")
```


# Try fixing issues if any 

```{r, warning=FALSE, echo=FALSE }
                                                                                             #===============================================================================
# Step 2: Swap REF/ALT and Genotypes
# ==============================================================================
message("\nStep 2: Identifying variants to swap...")

vcf_alt_alleles <- vcf_fix[, 'ALT']
fasta_at_alt_pos <- mapply(get_fasta_base,
                           chrom = vcf_fix[, "CHROM"],
                           pos = as.numeric(vcf_fix[, "POS"]),
                           MoreArgs = list(fasta_sequence = fasta_sequence))

if (all(is.na(fasta_at_alt_pos))) {
    stop("Failed to get FASTA reference for any ALT alleles. Stopping.")
}

idx_to_swap <- which(vcf_alt_alleles == fasta_at_alt_pos)

if (length(idx_to_swap) > 0) {
    message(paste0("Found ", length(idx_to_swap), " variants to swap. Swapping REF/ALT and genotypes..."))

    swapped_variants_table <- data.frame(
        CHROM = vcf_fix[idx_to_swap, 'CHROM'],
        POS = vcf_fix[idx_to_swap, 'POS'],
        Original_REF = vcf_fix[idx_to_swap, 'REF'],
        Original_ALT = vcf_alt_alleles[idx_to_swap]
    )

    original_ref_to_swap <- vcf@fix[idx_to_swap, 'REF']
    original_alt_to_swap <- vcf@fix[idx_to_swap, 'ALT']

    vcf@fix[idx_to_swap, 'REF'] <- original_alt_to_swap
    vcf@fix[idx_to_swap, 'ALT'] <- original_ref_to_swap

    genotypes <- vcf@gt
    if ("GT" %in% colnames(genotypes)) {
        for (i in idx_to_swap) {
            current_gt <- genotypes[i, "GT"]
            swapped_gt <- stringr::str_replace_all(current_gt, "0", "temp_0")
            swapped_gt <- stringr::str_replace_all(swapped_gt, "1", "0")
            swapped_gt <- stringr::str_replace_all(swapped_gt, "temp_0", "1")
            vcf@gt[i, "GT"] <- swapped_gt
        }
        message("REF/ALT and genotypes swapped successfully.")
    }

    message("Table of swapped variants:")
    print(swapped_variants_table)

} else {
    message("No variants found where ALT matches FASTA. No swaps performed.")
}

```



```{r, warning=FALSE, echo=FALSE }

#Comparison to VCFs genotyping arrays definitions - this is an example and needs adaptations


#if genotyping array known, check if variants are identical 
# e.g. here assuming the VCF was a MEGA Array - Problem is that exact version has to be known

#mega<-merge(ref_comparison_df, microArrayMega, by.y="V3", by.x="POS", all.y=TRUE) 
#mega$found<-ifelse(is.na(mega$VCF_REF), 0,1)

#ggplot(mega, aes(x=POS, y=found)) + geom_point(aes(color=found)) + ggtitle("Variants present on MEG-Array (1=yes, 0=not found)")

#ggplot(mega, aes(x = POS, y = found, color = factor(found))) +  geom_point(position = position_jitter(height = 0.1, seed = 123)) +  scale_color_manual(values = c("0" = "red", "1" = "blue"),  labels = c("0" = "Not Found", "1" = "Found")) +   labs(title = "Variants Present on MEG-Array", x = "Position", y = "Found (0 = No, 1 = Yes)", color = "Variant Status") +   theme_minimal() # A clean theme

```


# Prepare Haplogrep files + run HaploGrep2 

```{r, warning=FALSE, echo=FALSE}


# ==============================================================================
# Step 3: Update Haplogrep File
# ==============================================================================
message("\nStep 3: Preparing and updating Haplogrep file...")

final_pos <- vcf@fix[, 'POS']
concatenated_positions <- paste(final_pos, collapse = ";")
message(paste("Generated position string:", substr(concatenated_positions, 1, 100), "..."))

hsd_data <- read.delim(haplogrep_input_file, sep = "\t", stringsAsFactors = FALSE)
hsd_data[[3]]<-""
hsd_data[[2]] <- concatenated_positions

hsd_data <- hsd_data[, colSums(is.na(hsd_data)) == 0]

message("Haplogrep data updated. First few rows:")

write.table(hsd_data, file = output_haplogrep_file, sep = "\t", row.names = FALSE, col.names = FALSE, quote = FALSE)

message(paste("\nProcessing complete. Modified Haplogrep file saved to:", output_haplogrep_file))




# ==============================================================================
# Step 4: run Haplogrep File
# ==============================================================================

# Detect OS type
is_windows <- .Platform$OS.type == "windows"

# Use file.path() to build platform-safe paths
jar_path <- ("../../bin/haplogrep.jar")
#input_file <- "data/example.hsd"
input_file <- output_haplogrep_file
output_file <- "haplogroups.txt"

# Optional: Normalize paths (for more complex cases)
jar_path <- normalizePath(jar_path, winslash = "/", mustWork = FALSE)
input_file <- normalizePath(input_file, winslash = "/", mustWork = FALSE)
output_file <- normalizePath(output_file, winslash = "/", mustWork = FALSE)

# Build command safely
cmd <- paste(
  "java -jar", shQuote(jar_path), " classify ",
  "--in", shQuote(input_file),
  "--format hsd",
  "--phylotree", phylotreeVersion,
  "--out", shQuote(output_file)
)

# Run the command
system(cmd)
message("HaploGrep2 classify run")

# Read output if it exists
if (file.exists(output_file)) {
  haplogroups <- read.delim(output_file, sep = "\t", header = TRUE)
  haplogroupsList<-haplogroups %>%
    select(SampleID, Haplogroup)
}

  colnames(haplogroupsList) <- as.character(c("hg1", "hg2"))
  write.table(haplogroupsList, file = haplogroup_distcheck_file , sep = ";", row.names = FALSE, col.names = TRUE, quote = FALSE)
  

   
# Build command safely
cmd <- paste(
  "java -jar", shQuote(jar_path), " distance ",
  "--in", shQuote(haplogroup_distcheck_file),
  "--out", shQuote(haplogroup_distcheck_result_file)
)

# Run the command
system(cmd)
message("HaploGrep2 classify run")



# ==============================================================================
# Step 5: Haplogrep distance Analysis
# ==============================================================================


# Read output if it exists
if (file.exists(haplogroup_distcheck_result_file)) {
  haplogroupsDiff <- read.delim(haplogroup_distcheck_result_file, sep = ";", header = TRUE)
  #knitr::kable(haplogroupsDiff)
  }


haplogroupsDiff$hg1_Hgsubgroup<-str_extract(haplogroupsDiff$hg1, "[A-z]{1,4}+\\d*")
haplogroupsDiff$hg1_oneletter<-str_extract(haplogroupsDiff$hg1, "[A-z]*")

haplogroupsDiff$hg2_Hgsubgroup<-str_extract(haplogroupsDiff$hg2, "[A-z]{1,4}+\\d*")
haplogroupsDiff$hg2_oneletter<-str_extract(haplogroupsDiff$hg2, "[A-z]*")

haplogroupsDiff<-get_Supergroups(haplogroupsDiff)


sorted_haplogroups_diff <- haplogroupsDiff %>%
  arrange(desc(distance))

  #knitr::kable(head(sorted_haplogroups_diff, n = 10)) # Display top 10 rows (adjust 'n' as desired)



# Prepare data: Count flows between hg1 and hg2 MacroPhylo, handling NAs
plot_data_prep <- haplogroupsDiff %>%
  mutate(
    # Replace NAs with "Unknown" for plotting
    MacroPhylo_hg1 = ifelse(is.na(hg1_MacroPhylo), "Unknown", hg1_MacroPhylo),
    MacroPhylo_hg2 = ifelse(is.na(hg2_MacroPhylo), "Unknown", hg2_MacroPhylo)
  ) %>%
  # Count the occurrences of each unique flow
  group_by(MacroPhylo_hg1, MacroPhylo_hg2) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  ungroup()

#print(head(plot_data_prep))

# 1. Create distinct names for source and target nodes by prefixing
source_node_names_prefixed <- paste0("HG1: ", unique(plot_data_prep$MacroPhylo_hg1))
target_node_names_prefixed <- paste0("HG2: ", unique(plot_data_prep$MacroPhylo_hg2))

# 2. Combine all unique prefixed names to create the full list of nodes
all_node_names <- unique(c(source_node_names_prefixed, target_node_names_prefixed))

# 3. Create the nodes dataframe with 0-indexed IDs and original group for coloring
nodes_df <- data.frame(
  name = all_node_names,
  group = stringr::str_remove(all_node_names, "^HG[12]: ") # Extract original group for coloring
)
nodes_df$id <- 0:(nrow(nodes_df) - 1) # networkD3 requires 0-indexed IDs

# 4. Create the links dataframe using the prefixed names and their IDs
links_df <- plot_data_prep %>%
  # Map MacroPhylo_hg1 to its prefixed name
  mutate(Source_Prefixed = paste0("HG1: ", MacroPhylo_hg1),
         Target_Prefixed = paste0("HG2: ", MacroPhylo_hg2)) %>%
  # Join with nodes_df to get the 0-indexed IDs
  left_join(nodes_df, by = c("Source_Prefixed" = "name")) %>%
  rename(source_id = id) %>%
  left_join(nodes_df, by = c("Target_Prefixed" = "name")) %>%
  rename(target_id = id) %>%
  # Select the required columns for networkD3
  select(source = source_id, target = target_id, value = Count)

# Ensure no NAs in source or target IDs
if (any(is.na(links_df$source)) || any(is.na(links_df$target))) {
  warning("NA values found in source or target IDs after mapping. Check data preparation.")
  links_df <- links_df %>% filter(!is.na(source), !is.na(target))
}

# 5. Create the Sankey plot
sankeyNetwork(Links = links_df, Nodes = nodes_df,
              Source = "source", Target = "target",
              Value = "value", NodeID = "name",
              NodeGroup = "group", # Use the 'group' column for coloring
              units = "samples", fontSize = 12, nodeWidth = 30,
              colourScale = JS('d3.scaleOrdinal(d3.schemeCategory10);')) # Use a D3 color scheme


```

```{r,  warning=FALSE, echo=FALSE }


# Run the function for GSA-v3
gsa_results <- process_haplogroup_data(
  array_type = "GSA-24v3-0_A1",
  array_file_path = array_gsa_file,
  haplogrep_input_file = haplogrep_input_file,
  output_base_name = "gsa_output"
)


# Run the function for GSA-v4
#gsa4_results <- process_haplogroup_data(
#  array_type = "GSA-PGx-48v4",
#  array_file_path = array_gsa_file,
#  haplogrep_input_file = haplogrep_input_file,
#  output_base_name = "gsa4_output"
#)



# Run the function for Omni
omni_results <- process_haplogroup_data(
  array_type = "InfiniumOmni2-5-8v1-5_A1",
  array_file_path = array_omni_file,
  haplogrep_input_file = haplogrep_input_file,
  output_base_name = "omni_output"
)

# Run the function for Mega

mega_results <- process_haplogroup_data(
  array_type = "Multi-EthnicGlobal_D",
  array_file_path = array_mega_file,
  haplogrep_input_file = haplogrep_input_file,
  output_base_name = "mega_output"
)

#mega_A1_results <- process_haplogroup_data(
#  array_type = "Multi-EthnicGlobal_A1",
#  array_file_path = array_mega_A1_file,
#  haplogrep_input_file = haplogrep_input_file,
#  output_base_name = "mega_A1_output"
#)

#mega_D1_results <- process_haplogroup_data(
#  array_type = "Multi-EthnicGlobal_D1",
#  array_file_path = array_mega_D1_file,
#  haplogrep_input_file = haplogrep_input_file,
#  output_base_name = "mega_D1_output"
#)


axiom_ukb_results <- process_haplogroup_data(
  array_type = "Axiom UKB",
  array_file_path = array_axiom_ukb_file,
  haplogrep_input_file = haplogrep_input_file,
  output_base_name = "axiom_ukb_output"
)



# Initialize an empty list to store distance data
all_distances_list <- list()

# Add GSA data if available
if (!is.null(gsa_results) && !is.null(gsa_results$haplogroupsDiff)) {
  gsa_distances <- gsa_results$haplogroupsDiff %>%
    mutate(ArrayType = "GSA-24v3-0_A1") %>%
    select(ArrayType, distance) # Select only ArrayType and the Distance column
  all_distances_list[[length(all_distances_list) + 1]] <- gsa_distances
}

# Add GSA4 data if available
#if (!is.null(gsa4_results) && !is.null(gsa4_results$haplogroupsDiff)) {
#  gsa4_distances <- gsa4_results$haplogroupsDiff %>%
#    mutate(ArrayType = "GSA-PGx-48v4") %>%
#    select(ArrayType, distance) # Select only ArrayType and the Distance column
#  all_distances_list[[length(all_distances_list) + 1]] <- gsa4_distances
#}


# Add Omni data if available
if (!is.null(omni_results) && !is.null(omni_results$haplogroupsDiff)) {
  omni_distances <- omni_results$haplogroupsDiff %>%
    mutate(ArrayType = "InfiniumOmni2-5-8v1-5_A1") %>%
    select(ArrayType, distance)
  all_distances_list[[length(all_distances_list) + 1]] <- omni_distances
}

# Add Mega data if available
if (!is.null(mega_results) && !is.null(mega_results$haplogroupsDiff)) {
  mega_distances <- mega_results$haplogroupsDiff %>%
    mutate(ArrayType = "Multi-EthnicGlobal_D") %>%
    select(ArrayType, distance)
  all_distances_list[[length(all_distances_list) + 1]] <- mega_distances
}

# Add Mega A1 strand file from Rayner data if available
#if (!is.null(mega_A1_results) && !is.null(mega_A1_results$haplogroupsDiff)) {
#  mega_A1_distances <- mega_A1_results$haplogroupsDiff %>%
#    mutate(ArrayType = "Multi-EthnicGlobal_A1") %>%
#    select(ArrayType, distance)
#  all_distances_list[[length(all_distances_list) + 1]] <- mega_A1_distances
#}

# Add Mega D1 strand file from Rayner data if available
#if (!is.null(mega_D1_results) && !is.null(mega_D1_results$haplogroupsDiff)) {
#  mega_D1_distances <- mega_D1_results$haplogroupsDiff %>%
#    mutate(ArrayType = "Multi-EthnicGlobal_D1") %>%
#    select(ArrayType, distance)
#  all_distances_list[[length(all_distances_list) + 1]] <- mega_D1_distances
#}


# Add Axiom UKB data if available
if (!is.null(axiom_ukb_results) && !is.null(axiom_ukb_results$haplogroupsDiff)) {
  axiom_ukb_distances <- axiom_ukb_results$haplogroupsDiff %>%
    mutate(ArrayType = "Axiom-UKB") %>%
    select(ArrayType, distance)
  all_distances_list[[length(all_distances_list) + 1]] <- axiom_ukb_distances
}



# Combine all into a single data frame
combined_distances_df <- bind_rows(all_distances_list)

haplogroupsDiff$ArrayType="VCF"
diffVCF<-haplogroupsDiff %>%
  select(ArrayType, distance)

 
combined<-rbind(combined_distances_df, diffVCF)


if (!is.null(combined) && nrow(combined) > 0) {
  ggplot(combined, aes(x = ArrayType, y = distance, fill = ArrayType)) +
    geom_violin(trim = FALSE, alpha = 0.7) + # trim=FALSE ensures tails extend to min/max
    labs(
      title = "Comparison of Haplogroup Distances by Array Type",
      subtitle = "Violin Plot with Jittered Data Points",
      x = "Array Type",
      y = "Haplogroup Distance"
    ) +
    theme_minimal() +
    theme(legend.position = "none") # Remove legend if fill is only for X-axis
} else {
  cat("No distance data available for comparison.")
}

counted <- combined %>%
  group_by(ArrayType, distance) %>%
  summarise(count = n(), .groups = 'drop')



ggplot(combined, aes(x = distance, color = ArrayType)) +
  stat_ecdf(aes(linetype = (ArrayType == "VCF")), geom = "line", linewidth = 1) + 
  stat_ecdf(geom = "point", size = 1) +
  scale_x_continuous(breaks = 0:25) +
  scale_linetype_manual(values = c("TRUE" = "dashed", "FALSE" = "solid")) +
  guides(linetype = "none") + 
  theme_minimal(base_size = 12) +  coord_cartesian(xlim = c(1, max(combined$distance, na.rm = TRUE)))
```


## Haplogroup Differences Table (GSA)

```{r gsa-table, echo=FALSE, results='asis'}
if (!is.null(gsa_results) && !is.null(gsa_results$haplogroupsDiff)) {
   sorted_haplogroups_diff <- gsa_results$haplogroupsDiff %>%
   arrange(desc(distance))

 # knitr::kable(head(sorted_haplogroups_diff, n = 10)) # Display top 10 rows (adjust 'n' as desired)

} else {
  cat("No haplogroup difference data available for GSA.")
}

if (!is.null(gsa_results) && !is.null(gsa_results$sankey_plot)) {
  gsa_results$sankey_plot
} else {
  cat("No Sankey plot available for GSA.")
}
```

## Omni Haplogroup Analysis

### Haplogroup Differences Table (Omni)
```{r omni-table, echo=FALSE, results='asis'}
if (!is.null(omni_results) && !is.null(omni_results$haplogroupsDiff)) {
    sorted_haplogroups_diff <- omni_results$haplogroupsDiff %>%
    arrange(desc(distance))

  #knitr::kable(head(sorted_haplogroups_diff, n = 10)) # Display top 10 rows (adjust 'n' as desired)

} else {
  cat("No haplogroup difference data available for Omni.")
}

if (!is.null(omni_results) && !is.null(omni_results$sankey_plot)) {
  omni_results$sankey_plot
} else {
  cat("No Sankey plot available for Omni.")
}



```

## Mega Haplogroup Analysis

### Haplogroup Differences Table (Mega)
```{r mega-table, echo=FALSE, results='asis'}
if (!is.null(mega_results) && !is.null(mega_results$haplogroupsDiff)) {
    sorted_haplogroups_diff <- mega_results$haplogroupsDiff %>%
    arrange(desc(distance))

  #knitr::kable(head(sorted_haplogroups_diff, n = 10)) # Display top 10 rows (adjust 'n' as desired)

} else {
  cat("No haplogroup difference data available for Mega.")
}

if (!is.null(mega_results) && !is.null(mega_results$sankey_plot)) {
  mega_results$sankey_plot
} else {
  cat("No Sankey plot available for Mega.")
}

```

## Axiom UKB Haplogroup Analysis

### Haplogroup Differences Table (Axiom UKB)
```{r axiom-ukb-table, echo=FALSE, results='asis'}
if (!is.null(axiom_ukb_results) && !is.null(axiom_ukb_results$haplogroupsDiff)) {
    sorted_haplogroups_diff <- axiom_ukb_results$haplogroupsDiff %>%
    arrange(desc(distance))

 # knitr::kable(head(sorted_haplogroups_diff, n = 10)) # Display top 10 rows (adjust 'n' as desired)

} else {
  cat("No haplogroup difference data available for Omni.")
}

if (!is.null(axiom_ukb_results) && !is.null(axiom_ukb_results$sankey_plot)) {
  axiom_ukb_results$sankey_plot
} else {
  cat("No Sankey plot available for Omni.")
}



```


# Compare Macro - Haplogroups

### Prepare Data for MacroPhylo Comparison
```{r combine-macrophylos-for-plot, echo=FALSE}
# Initialize an empty list to store MacroPhylo data
all_macrophylos_list <- list()

# Function to safely extract and label MacroPhylo data
extract_macrophylos <- function(results_object, array_label) {
  if (!is.null(results_object) && !is.null(results_object$haplogroupsDiff)) {
    df <- results_object$haplogroupsDiff %>%
      mutate(
        ArrayType = array_label,
        # Convert NA MacroPhylo values to "Unknown" for consistent plotting
        hg2_MacroPhylo_Clean = ifelse(is.na(hg2_MacroPhylo), "Unknown", hg2_MacroPhylo)
      ) %>%
      select(ArrayType, hg2_MacroPhylo_Clean)
    return(df)
  } else {
    return(NULL)
  }
}

# Add GSA data
all_macrophylos_list[[length(all_macrophylos_list) + 1]] <- extract_macrophylos(gsa_results, "GSA")

# Add GSA4 data
#all_macrophylos_list[[length(all_macrophylos_list) + 1]] <- #extract_macrophylos(gsa4_results, "GSA4")

# Add Omni data
all_macrophylos_list[[length(all_macrophylos_list) + 1]] <- extract_macrophylos(omni_results, "Omni")

# Add Mega data
all_macrophylos_list[[length(all_macrophylos_list) + 1]] <- extract_macrophylos(mega_results, "Mega")

# Add Axiom UKB data
all_macrophylos_list[[length(all_macrophylos_list) + 1]] <- extract_macrophylos(axiom_ukb_results, "Axiom UKB")


# -------------------------------------------------------------
# ADDITION: Incorporate additional dataframe's MacroPhylo data
# -------------------------------------------------------------

# --- Configuration for additional dataframe (same as before) ---
my_additional_dataframe_name <- "haplogroupsDiff" 

column_with_macrophylos_in_additional_df <- "hg2_MacroPhylo" # REPLACE if your column is different
new_label_for_additional_data <- "VCF" # REPLACE with your label

if (exists(my_additional_dataframe_name) && is.data.frame(get(my_additional_dataframe_name))) {
  temp_additional_df <- get(my_additional_dataframe_name)

  if (column_with_macrophylos_in_additional_df %in% colnames(temp_additional_df)) {
    additional_macrophylos_data <- temp_additional_df %>%
      mutate(
        ArrayType = new_label_for_additional_data,
        hg2_MacroPhylo_Clean = ifelse(is.na(!!sym(column_with_macrophylos_in_additional_df)), "Unknown", !!sym(column_with_macrophylos_in_additional_df))
      ) %>%
      select(ArrayType, hg2_MacroPhylo_Clean)

    all_macrophylos_list[[length(all_macrophylos_list) + 1]] <- additional_macrophylos_data
    message(paste("Successfully added MacroPhylo data from '", my_additional_dataframe_name, "' with label '", new_label_for_additional_data, "'", sep=""))

  } else {
    warning(paste("MacroPhylo column '", column_with_macrophylos_in_additional_df, "' not found in dataframe '", my_additional_dataframe_name, "'. Skipping addition of MacroPhylo data.", sep=""))
  }
} else {
  warning(paste("Dataframe '", my_additional_dataframe_name, "' not found for MacroPhylo comparison. Skipping addition.", sep=""))
}

# -------------------------------------------------------------
# End of Addition
# -------------------------------------------------------------

expected <- haplogroupsDiff %>%
  transmute(
    ArrayType = "Expected",
    hg2_MacroPhylo_Clean = ifelse(is.na(hg1_MacroPhylo), "Unknown", hg1_MacroPhylo)
  )


expected_df <- expected %>%
  group_by(ArrayType, hg2_MacroPhylo_Clean) %>%
  summarise(Count=n()) %>%
  mutate(Expected_Count = Count) %>%
  select(hg2_MacroPhylo_Clean, Expected_Count)


combined_macrophylos_df <- bind_rows(all_macrophylos_list[!sapply(all_macrophylos_list, is.null)])


# Optional: Set factor levels for ArrayType and hg2_MacroPhylo_Clean for consistent plotting order
if (!is.null(combined_macrophylos_df) && nrow(combined_macrophylos_df) > 0) {
  # Order ArrayTypes (e.g., GSA, Omni, Mega, External)
  combined_macrophylos_df$ArrayType <- factor(
    combined_macrophylos_df$ArrayType,
    levels = c("GSA", "Omni", "Mega", "Axiom UKB", new_label_for_additional_data, "Expected")
  )
}

```

### Comparison of MacroPhylo Distributions

```{r, echo=FALSE, out.width="80%"}
if (!is.null(combined_macrophylos_df) && nrow(combined_macrophylos_df) > 0) {
  # Calculate proportions for plotting
  plot_data_macrophylos <- combined_macrophylos_df %>%
    group_by(ArrayType, hg2_MacroPhylo_Clean) %>%
    summarise(Count = n(), .groups = 'drop_last') %>%
    mutate(Proportion = Count / sum(Count)) %>%
    ungroup()
  
  
  combined_df <- plot_data_macrophylos %>%
  left_join(expected_df, by = "hg2_MacroPhylo_Clean") %>%
  rename(Observed_Count = Count) %>%
  tidyr::pivot_longer(
    cols = c("Observed_Count", "Expected_Count"),
    names_to = "Type",
    values_to = "Count"
  )
  
combined_df$Type <- recode(combined_df$Type,
                        "Observed_Count" = "Observed",
                        "Expected_Count" = "Expected")

combined_df <- combined_df %>%
  rename(ArrayType = ArrayType.x) %>%
  select(-ArrayType.y)  


p1<-ggplot(combined_df, aes(x = hg2_MacroPhylo_Clean, y = Count, fill = Type)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  facet_wrap(~ ArrayType, scales = "free_y", ncol=1) +
  scale_fill_manual(values = c("Observed" = "#1f77b4", "Expected" = "grey70")) +
  labs(
    title = "Observed vs Expected Counts of Macro-Haplogroups per ArrayType",
    x = "Macro Haplogroup",
    y = "Count",
    fill = "Type"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.major.x = element_blank()
  )  + theme(legend.position="bottom") 

  print(p1)

ggplot(plot_data_macrophylos, aes(x = hg2_MacroPhylo_Clean, y = Proportion, fill = hg2_MacroPhylo_Clean)) + # X is now MacroPhylo, fill by MacroPhylo
    geom_col(width = 0.8) +
    facet_wrap(~ ArrayType,     # Facet by ArrayType (your methods)
               ncol = 1,     # Let ggplot decide number of columns/rows, or set a specific number (e.g., ncol = 2)
               scales = "free_y" # Allow y-axis to vary for each facet, as proportions might differ greatly between methods
    ) +
    labs(
      title = "Distribution of Macro Haplogroups per MicroArray Positions",
      x = "Macro Haplogroup",
      y = "Proportion",
      fill = "Macro Haplogroup" # Legend title
    ) +
    scale_y_continuous(labels = scales::percent) + # Format y-axis as percentage
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1), # Rotate x-axis labels (Macro Haplogroup names)
      strip.background = element_rect(fill = "grey90", color = "grey90"), # Background for facet labels
      strip.text = element_text(face = "bold") # Bold facet labels
    )  + theme(legend.position="none") 
  
} else {
  cat("No Macro Haplogroup data available for comparison.")
}

```
