---
title: "Microarray Performance Across Populations"
author: "Hansi"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
params:
  data_dir: "./"
  files_AUC: "AUC_results_.*\\.txt$"
  files_HaploDiff: "haplogroups_diff_all_.*\\.txt$"
  pareto_w: 0.25
  score_w: 1.0
  seed: 123
  w_auc: 0.6
  w_match: 0.4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 10, fig.height = 6)

# Load libraries
library(tidyverse)
library(janitor)
library(ggrepel)
library(scales)
library(ComplexHeatmap)
library(kableExtra)
library(DT) # For interactive tables

set.seed(params$seed)
```

# 1. Summary: Top 10 Overall

This table ranks arrays by their average Z-score performance across all detected populations.

```{r, warning=FALSE, echo=FALSE}
# 1. Load and Harmonize
files_AUC <- list.files(params$data_dir, pattern = params$files_AUC, full.names = TRUE)
stopifnot(length(files_AUC) > 0)

df_raw <- map_df(files_AUC, ~read_csv(.x, show_col_types = FALSE) %>% clean_names()) %>%
  mutate(population = coalesce(!!!select(., matches("^pop(ulation)?(_\\d+)?$")))) %>%
  select(array_type, population, auc_normalized, mean_distance, mean_match_score, tier) %>%
  mutate(across(c(auc_normalized, mean_distance, mean_match_score), as.numeric)) %>%
  filter(!is.na(population), !is.na(auc_normalized))

df_AUC_ALL<-subset(df_raw, population=="ALL") 
df_AUC<-subset(df_raw, population!="ALL")



# 2. Calculate Metrics weighting 

safe_z <- function(x) {
  m <- mean(x, na.rm = TRUE)
  s <- stats::sd(x, na.rm = TRUE)
  if (is.na(s) || s == 0) return(rep(0, length(x)))  # if no variation, set z = 0
  (x - m) / s
}

df_scored <- df_AUC %>%
  group_by(population) %>%
  mutate(
    auc_z    = safe_z(auc_normalized),    # larger is better
    dist_z   = safe_z(mean_distance),     # smaller is better 
    match_z  = safe_z(mean_match_score),  # larger is better
    # Composite based on AUC and MATCH only (both larger is better)
    composite_z = params$w_auc * auc_z + params$w_match * match_z,
    pareto_score = auc_normalized - (params$pareto_w * mean_distance)
  ) %>% 
  mutate(composite_01 = pnorm(composite_z)) %>%
  ungroup()

# 3. Summary Table Logic
summary_tab <- df_scored %>%
  group_by(array_type) %>%
    summarize(
    `Avg AUC` = mean(auc_normalized, na.rm = TRUE),
    `Avg Distance` = mean(mean_distance, na.rm = TRUE),
    `Avg Match Score` = mean(mean_match_score, na.rm = TRUE),
    `Global Score (Z)` = mean(composite_z, na.rm = TRUE),
    `Pops` = n_distinct(population)
  ) %>%
  slice_max(order_by = `Global Score (Z)`, n = 10) %>%
  arrange(desc(`Global Score (Z)`)) %>%
  mutate(across(where(is.numeric), ~round(.x, 3)))

summary_tab <- summary_tab %>% rename(Array = array_type)


summary_tab %>%
  kable(caption = "Top 10 Microarrays (Ranked by Cross-Population Z-Score)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = F) %>%
  row_spec(0, bold = T, color = "white", background = "#2A9D8F")



```

# 2. Global Performance Visuals

AUC Distribution by Population

```{r, warning=FALSE, echo=FALSE}
ggplot(df_scored, aes(population, auc_normalized, fill = population)) +
  geom_violin(alpha = 0.4, color = NA) + 
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.8) +
  theme_minimal() + 
  scale_fill_viridis_d(option = "B") +
  labs(title = "Haplogroup Assignment Performance AUC", y = "Normalized AUC")

ggplot(df_scored, aes(population, mean_match_score, fill = population)) +
  geom_violin(alpha = 0.4, color = NA) +
  geom_boxplot(width = 0.1, outlier.shape = NA, alpha = 0.8) +
  theme_minimal() + 
  scale_fill_viridis_d(option = "B") +
  labs(title = "Mean Haplogroup Matching Score", y = "mean matching score")

```

Clustering: Array Profiles

```{r, warning=FALSE, echo=FALSE, fig.height=12, fig.width=9}
# Function to generate a standardized heatmap
make_heatmap <- function(metric_col, label) {
  mat <- df_scored %>%
    select(array_type, population, !!sym(metric_col)) %>%
    pivot_wider(names_from = population, values_from = !!sym(metric_col)) %>%
    column_to_rownames("array_type") %>% 
    as.matrix()
  
  ht=Heatmap(mat, name = label, heatmap_legend_param = list(direction = "horizontal"),
          col = colorRampPalette(c("#E63946","#F5C518","#2A9D8F"))(100),
          row_names_gp = gpar(fontsize = 7),
          column_title = paste("Arrays Clustered by", label))
  
  draw(ht, heatmap_legend_side = "bottom")
  
}

make_heatmap("auc_normalized", "AUC")

make_heatmap("mean_match_score", "Mean Match Score")

```

# 3. Population Breakdowns

```{r, warning=FALSE, echo=FALSE}
# Define reusable plotting logic per population
plot_population <- function(pop_name) {
  pop_data <- df_scored %>% filter(population == pop_name)
  
  p <- ggplot(pop_data, aes(mean_match_score, auc_normalized, color = tier)) +
    geom_point(alpha = 0.6, size = 2.5) +
    geom_text_repel(data = slice_max(pop_data, pareto_score, n = 5), 
                    aes(label = array_type), size = 3, fontface = "bold") +
    theme_light() +
    labs(title=pop_name, subtitle = paste("Top 5 Pareto performers labeled"),
         x = "Mean Haplogrouping Matching Score (Higher is Better)", 
         y = "AUC (Higher is Better)") +
    scale_color_brewer(palette = "Set1")
  
  return(p)
}
```

```{r, results='asis', echo=FALSE, message=FALSE, warning=FALSE}
all_pops <- sort(unique(trimws(df_scored$population)))

for (pop in all_pops) {
  p <- plot_population(pop)   # returns a ggplot; no side effects
  if (inherits(p, "ggplot")) print(p)  
}

```

# 4. Haplogroup Match Details

Overview of Haplogroup matching details a) perfect match, subhaplogroup match, one-letter match, macrhaplogroup match or mismatch (e.g. L vs H)

```{r, warning=FALSE, echo=FALSE}
# 1. Load and Harmonize
files_HaploDiff <- list.files(params$data_dir, pattern = params$files_HaploDiff, full.names = TRUE)
stopifnot(length(files_HaploDiff) > 0)

df_match <- files_HaploDiff %>%
  set_names() %>% 
  map_df(~read_delim(.x, show_col_types = FALSE), .id = "file_origin") %>%
  clean_names() %>%
  mutate(
    population = file_origin %>% 
      str_remove_all("\\.txt") %>% 
      str_remove_all(".//haplogroups_diff_all_") 
    ) %>%
  select(-file_origin)

binary_cols <- c("matched_haplogroup", "matched_subgroup", "matched_macro", "matched_oneletter")

summary <- df_match %>%
  group_by(population) %>%
  summarise(
    across(all_of(binary_cols), ~ mean(.x, na.rm = TRUE)),
    distance = mean(distance, na.rm = TRUE),
    .groups = "drop"
  )

long_bin <- summary %>%
  pivot_longer(cols = all_of(binary_cols), names_to = "metric", values_to = "proportion") %>%
  mutate(
    metric = recode(metric,
                    matched_haplogroup = "Perfect hits",
                    matched_subgroup   = "Subgroup concordance",
                    matched_macro      = "Macrohaplogroup concordance",
                    matched_oneletter  = "One-letter concordance")
  )

# Order populations by size (or choose another ordering)
long_bin <- long_bin %>%
  mutate(population = fct_reorder(population, -n_arrays))

summary <- summary %>%
  mutate(population = factor(population, levels = levels(long_bin$population)))


ggplot() +
  geom_col(
    data = long_bin,
    aes(x = population, y = proportion, fill = metric),
    position = position_dodge(width = 0.8),
    width = 0.75,
    color = "black"
  ) +
  geom_line(
    data = summary,
    aes(x = population, y = distance / scale_factor, group = 1),
    color = "red", linewidth = 1.2
  ) +
  geom_point(
    data = summary,
    aes(x = population, y = distance / scale_factor),
    color = "red", size = 2
  ) +
  scale_y_continuous(
    name = "Proportion (mean of 0/1 metrics)",
    limits = c(0, 1),
    sec.axis = sec_axis(~ . * scale_factor, name = "Mean node distance")
  ) +
  scale_fill_brewer(palette = "Set2", name = "Metric") +
  labs(x = "Population") +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

```

# 5. Data Explorer

Use the search box below to filter specific arrays. You can copy this data or use the buttons below to save the full scored dataset.

```{r, warning=FALSE, echo=FALSE}

df_ranked <- df_scored %>%
  group_by(population) %>%
  arrange(population, desc(auc_normalized), .by_group = TRUE) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  mutate(
    pop_rank_auc = min_rank(desc(auc_normalized)),   # ties share the same rank (1,1,3…)
  ) %>% 
  select(array_type, population, pop_rank_auc, auc_normalized, auc_z, mean_distance, mean_match_score, tier)  %>% 
  ungroup()

datatable(df_ranked, extensions = 'Buttons', filter = "top", options = list(
    dom = 'Bfrtip',
    buttons = c('copy', 'csv', 'excel')
  ))

# Optional: Save a physical CSV of processed scores to the directory
write_csv(df_scored, "processed_microarray_scores.csv")
```
