---
title: "Microarray Performance Across Populations"
author: "Hansi"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
params:
  data_dir: "./"      # directory with the six CSV files
  file_pattern: "AUC_results_.*\\.txt$"       # pattern to match input files
  top_n_overall: 15             # how many top arrays to label/plot
  pareto_w: 0.25                # weight of distance in pareto score (AUC - w * distance)
  score_w: 1.0                  # weight on distance in overall score: auc_z - w * dist_z
  seed: 123
editor_options: 
  markdown: 
    wrap: 72
---

```{r}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE, fig.width = 8, fig.height = 5
)
set.seed(params$seed)
```

# 1. Overview and data expectations

This report aggregates per-population TXT files containing microarray
performance metrics for haplogroup assignment. Expected columns
(case-insensitive, will be cleaned to snake_case): - ArrayType →
array_type - AUC_Normalized → auc_normalized (higher is better) -
Mean_Distance → mean_distance (lower is better) - n → n - Tier → tier -
Population → population

We will: - Ingest and harmonize the six TXT files - Diagnose common
schema issues (e.g., duplicated/nested population columns) - Compute
per-population ranks and z-scores - Build overall scores across
populations - Visualize performance distributions, Pareto trade-offs,
best arrays, heatmaps, and similarity maps

Z-score standardization within each population: $$
z = \frac{x - \mu}{\sigma}
$$ Correlation-based distance for similarity mapping (using AUC
profiles): $$
d_{ij} = 1 - r_{ij}
$$

# 2. Packages

```{r}
library(tidyverse)   # dplyr, ggplot2, tidyr, readr, forcats, etc.
library(janitor)     # clean_names
library(ggrepel)     # smart text labels
library(scales)      # number_format, percent
library(broom)       # tidy model outputs (optional)
```

# 3. Read and harmonize data

We robustly read all CSVs, clean column names, and fix cases where
population becomes a data-frame column (df-col) due to header
differences. We also ensure numeric types for measures and handle
duplicate "population-like" columns by coalescing.

```{r}
# Locate files
files <- list.files(params$data_dir, pattern = params$file_pattern, full.names = TRUE)
stopifnot(length(files) > 0)

# Helper to read one file with cleaned names
read_one <- function(path) {
  readr::read_csv(path, show_col_types = FALSE) |>
    janitor::clean_names()
}

# Read all
raw_list <- purrr::map(files, read_one)

# Inspect schemas (names only) to spot inconsistencies
schemas <- purrr::map(raw_list, names)
schemas

# Bind rows (may create df-cols if there are conflicting names)
df <- dplyr::bind_rows(raw_list)

# Ensure "population" exists or can be constructed from similarly named columns
pop_candidates <- df |> dplyr::select(dplyr::matches("^population(\\_\\d+)?$|^pop(ulation)?$"))
if (ncol(pop_candidates) == 0) {
  stop("No population column found. Please check headers in your CSV files.")
}

# If we have multiple population-like columns, coalesce them (first non-missing)
if (ncol(pop_candidates) > 1) {
  df <- df |>
    dplyr::mutate(population = dplyr::coalesce(!!!pop_candidates)) |>
    dplyr::select(-dplyr::matches("^population\\_\\d+$"))
} else {
  # If population is a df-col with one column, flatten it
  if (is.data.frame(df$population)) {
    inner <- df$population
    if (ncol(inner) == 1) {
      df$population <- inner[[1]]
    } else {
      df$population <- do.call(dplyr::coalesce, as.list(inner))
    }
  }
}

# Standardize column names to the expected set (renames noop if already correct)
df <- df |>
  dplyr::rename(
    array_type    = dplyr::any_of("array_type"),
    auc_normalized = dplyr::any_of("auc_normalized"),
    mean_distance  = dplyr::any_of("mean_distance"),
    n              = dplyr::any_of("n"),
    tier           = dplyr::any_of("tier"),
    population     = dplyr::any_of("population"),
    mean_match_score= dplyr::any_of("mean_match_score")
  )

# Enforce types
df <- df |>
  dplyr::mutate(
    array_type    = as.character(array_type),
    population    = as.character(population),
    tier          = as.character(tier),
    auc_normalized = suppressWarnings(as.numeric(auc_normalized)),
    mean_distance  = suppressWarnings(as.numeric(mean_distance)),
    n              = suppressWarnings(as.integer(n)),
    mean_match_score  = suppressWarnings(as.numeric(mean_match_score))
  )

# Basic sanity checks
stopifnot(all(c("array_type", "population", "auc_normalized", "mean_distance", "mean_match_score") %in% names(df)))
df <- df |>
  dplyr::filter(!is.na(array_type), !is.na(population)) |>
  dplyr::mutate(
    population = factor(population),
    array_type = factor(array_type)
  )

glimpse(df)
dplyr::count(df, population, name = "rows_per_population")
```

Notes: - If you still encounter "cannot xtfrm data frames," examine
names(df) and str(df\$population) to ensure it's a simple vector after
the steps above. - For emoji in tier, we keep it as character or factor;
both work with ggplot.

# 4. Quick summaries and derived metrics

We compute per-population ranks, z-scores, and an overall composite
score across populations.

-   Rank within population:
    -   Higher AUC = better rank
    -   Lower distance = better rank
-   Z-scores within population:
    -   Standardizes distributions across populations
-   Overall score across populations:
    -   Average of per-row composite: score_row = auc_z − w × dist_z
    -   Higher is better. w is adjustable via params\$score_w.

```{r}
# Rank within each population
df_ranked <- df %>%
  group_by(population) %>%
  mutate(
    auc_rank  = dense_rank(desc(auc_normalized)),
    dist_rank = dense_rank(mean_distance) # smaller distance = better
  ) %>%
  ungroup()

# Population-standardized z-scores
df_scaled <- df %>%
  group_by(population) %>%
  mutate(
    auc_z  = (auc_normalized - mean(auc_normalized, na.rm = TRUE)) /
             sd(auc_normalized, na.rm = TRUE),
    dist_z = (mean_distance  - mean(mean_distance,  na.rm = TRUE)) /
             sd(mean_distance,  na.rm = TRUE)
  ) %>%
  ungroup()

# Composite overall score per array across populations
w <- params$score_w
overall <- df_scaled %>%
  mutate(row_score = auc_z - w * dist_z) %>%
  group_by(array_type) %>%
  summarize(
    n_pops   = n_distinct(population),
    mean_auc = mean(auc_normalized, na.rm = TRUE),
    mean_dist= mean(mean_distance,  na.rm = TRUE),
    score    = mean(row_score,      na.rm = TRUE)
  ) %>%
  arrange(desc(score))

head(overall, 10)
```

# 5. AUC distribution by population

```{r}
ggplot(df, aes(population, auc_normalized, fill = population)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  scale_y_continuous(labels = number_format(accuracy = 0.001)) +
  labs(title = "AUC distribution by population",
       x = "Population", y = "AUC (normalized)") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 5.1 Mean Match Scores distribution by population

```{r}
ggplot(df, aes(population, mean_match_score, fill = population)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  scale_y_continuous(labels = number_format(accuracy = 0.001)) +
  labs(title = "Mean Haplogroup Matching distribution by population (higher=better)",
       x = "Population", y = "Mean Match Score") +
  theme_minimal() +
  theme(legend.position = "none")
```

## 5.2 Mean Haplogroup Distance distribution by population

```{r}
ggplot(df, aes(population, mean_distance, fill = population)) +
  geom_violin(trim = FALSE, alpha = 0.5) +
  geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.7) +
  scale_y_continuous(labels = number_format(accuracy = 0.001)) +
  labs(title = "Mean Haplogroup Distance by population (lower=better)",
       x = "Population", y = "Mean Match Score") +
  theme_minimal() +
  theme(legend.position = "none")
```

Interpretation: - Compare medians and spread of AUC across populations
to identify populations where haplogroup assignment is inherently harder
or easier.

# 6. AUC vs distance within each population (Pareto view)

We visualize the trade-off between AUC (higher is better) and mean
distance (lower is better). We label top performers by a simple
Pareto-like score: $$
\text{pareto} = \text{AUC} - w \times \text{distance}
$$ with adjustable weight w = params\$pareto_w.

```{r}
pareto_w <- params$pareto_w

top_per_pop <- df %>%
  group_by(population) %>%
  mutate(pareto = auc_normalized - pareto_w * mean_distance) %>%
  slice_max(order_by = pareto, n = 5, with_ties = FALSE) %>%
  ungroup()

ggplot(df, aes(mean_distance, auc_normalized, color = tier)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "gray40", linetype = 2) +
  geom_text_repel(
    data = top_per_pop,
    aes(label = array_type),
    size = 3, max.overlaps = 20, seed = params$seed
  ) +
  facet_wrap(~ population) +
  labs(title = "AUC vs Mean Distance per population",
       x = "Mean distance (lower is better)",
       y = "AUC (higher is better)",
       color = "Tier") +
  theme_minimal()
```

# 7. Top microarrays overall (composite score)

```{r}
top_overall <- overall %>%
  slice_max(order_by = score, n = params$top_n_overall) %>%
  mutate(array_type = fct_reorder(as.character(array_type), score))

ggplot(top_overall, aes(array_type, score)) +
  geom_col(fill = "#2b8cbe") +
  coord_flip() +
  labs(title = paste0("Top microarrays overall (composite z-score: AUC_z - ",
                      w, " * Distance_z)"),
       x = "Array type", y = "Overall score") +
  theme_minimal()
```

# 8. Heatmap: performance by population per array (AUC)

We cluster arrays based on their AUC profiles across populations and
visualize a heatmap.

```{r}
# Wide AUC matrix
auc_wide <- df %>%
  select(array_type, population, auc_normalized) %>%
  mutate(array_type = as.character(array_type)) %>%
  pivot_wider(names_from = population, values_from = auc_normalized)

auc_mat <- auc_wide %>%
  column_to_rownames("array_type") %>%
  as.matrix()

# Impute NAs per population with column mean
for (j in seq_len(ncol(auc_mat))) {
  nas <- is.na(auc_mat[, j])
  if (any(nas)) auc_mat[nas, j] <- mean(auc_mat[, j], na.rm = TRUE)
}

# Cluster arrays by their AUC profiles (scaled columns)
auc_mat_scaled <- scale(auc_mat, center = TRUE, scale = TRUE)
arr_order <- hclust(dist(auc_mat_scaled))$order
array_levels <- rownames(auc_mat_scaled)[arr_order]

heat_df <- auc_wide %>%
  pivot_longer(-array_type, names_to = "population", values_to = "auc") %>%
  mutate(
    array_type = factor(array_type, levels = array_levels),
    population = factor(population, levels = sort(unique(population)))
  )

ggplot(heat_df, aes(population, array_type, fill = auc)) +
  geom_tile() +
  scale_fill_viridis_c(option = "C", name = "AUC") +
  labs(title = "AUC heatmap by population (arrays clustered by AUC profile)",
       x = "Population", y = "Array type") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))
```

# 9. Similar behavior via PCA (AUC + distance features)

We build features combining AUC and distance across populations and run
PCA to visualize arrays with similar cross-population behavior.

```{r}
feat_auc <- df %>%
  select(array_type, population, auc_normalized) %>%
  pivot_wider(names_from = population, values_from = auc_normalized, names_prefix = "auc_")

feat_dist <- df %>%
  select(array_type, population, mean_distance) %>%
  pivot_wider(names_from = population, values_from = mean_distance, names_prefix = "dist_")

feat <- feat_auc %>%
  left_join(feat_dist, by = "array_type") %>%
  column_to_rownames("array_type") %>%
  as.data.frame()

# Impute missing values with column means
for (j in seq_len(ncol(feat))) {
  nas <- is.na(feat[, j])
  if (any(nas)) feat[nas, j] <- mean(feat[, j], na.rm = TRUE)
}

# PCA on scaled features
pca <- prcomp(scale(feat), center = TRUE, scale. = TRUE)

scores <- as_tibble(pca$x, rownames = "array_type") %>%
  left_join(overall, by = "array_type")

ggplot(scores, aes(PC1, PC2, color = score)) +
  geom_point(size = 2, alpha = 0.8) +
  scale_color_viridis_c(option = "C") +
  ggrepel::geom_text_repel(
    aes(label = ifelse(rank(-score) <= params$top_n_overall, array_type, "")),
    size = 3, max.overlaps = 20, seed = params$seed
  ) +
  labs(title = "PCA of microarrays across populations (AUC + distance features)",
       x = paste0("PC1 (", scales::percent(summary(pca)$importance[2, 1]), ")"),
       y = paste0("PC2 (", scales::percent(summary(pca)$importance[2, 2]), ")"),
       color = "Overall score") +
  theme_minimal()
```

# 10. Optional: similarity map via correlation of AUC profiles

We compute pairwise similarity using Pearson correlation across
population AUC profiles, then project into 2D with classical MDS.

The distance is: $$
d_{ij} = 1 - r_{ij}
$$

```{r}
auc_prof <- df %>%
  select(array_type, population, auc_normalized) %>%
  pivot_wider(names_from = population, values_from = auc_normalized) %>%
  column_to_rownames("array_type") %>%
  as.matrix()

# Impute NAs by column means
for (j in seq_len(ncol(auc_prof))) {
  nas <- is.na(auc_prof[, j])
  if (any(nas)) auc_prof[nas, j] <- mean(auc_prof[, j], na.rm = TRUE)
}

# Correlation-based distance
cor_mat <- cor(t(auc_prof), use = "pairwise.complete.obs")
dist_cor <- as.dist(1 - cor_mat)

# MDS embedding
mds <- cmdscale(dist_cor, k = 2)
mds_df <- as_tibble(mds, .name_repair = "minimal") %>%
  set_names(c("x", "y")) %>%
  mutate(array_type = rownames(mds)) %>%
  left_join(overall, by = "array_type")

ggplot(mds_df, aes(x, y, color = score)) +
  geom_point(size = 2, alpha = 0.9) +
  scale_color_viridis_c(option = "C") +
  ggrepel::geom_text_repel(
    aes(label = ifelse(rank(-score) <= params$top_n_overall, array_type, "")),
    size = 3, max.overlaps = 20, seed = params$seed
  ) +
  labs(title = "Similarity map (MDS on 1 - correlation of AUC profiles)",
       x = "MDS1", y = "MDS2", color = "Overall score") +
  theme_minimal()
```

# 11. Interpretation tips

-   AUC distribution plot: Identify populations with lower central
    tendency or broader spread, which may indicate greater difficulty
    assigning haplogroups.
-   Pareto plots: Within each population, top-left points (high AUC, low
    distance) are best. The LOESS curve shows the average trade-off.
-   Top overall bar chart: Highlights arrays that are robust across
    populations given the composite score. Adjust the distance weight if
    needed.
-   Heatmap: Arrays with consistently high AUC across all populations
    are desirable; clustering groups arrays with similar performance
    profiles.
-   PCA/MDS: Nearby points behave similarly across populations. Use
    these to find substitutes or arrays that cluster with known strong
    performers.

# 12. Reproducibility and diagnostics

-   Inspect input schemas:

``` r
unique(schemas)
```

-   If issues persist with the population column:

``` r
str(df$population)
names(df)
```

-   Validate complete cases per array-population:

``` r
df %>% count(array_type, population) %>% count(array_type) %>% arrange(n)
```

-   Save intermediate outputs:

``` r
readr::write_csv(overall, "overall_scores.csv")
```

-   Session info:

``` r
sessionInfo()
```

Appendix: Parameter notes - pareto_w controls the trade-off in labeling
top performers in the faceted scatter. Larger values penalize distance
more strongly. - score_w controls the contribution of distance in the
overall composite: score_row = auc_z − w × dist_z. - You may set both to
0 to focus purely on AUC, or increase them to emphasize distance.

Formulas used - Per-population z-score for metric x: $$
z = \frac{x - \mu}{\sigma}
$$ - Correlation-based distance between arrays i and j: $$
d_{ij} = 1 - r_{ij}
$$

If you share a few representative rows from each file (including
headers), we can further specialize the data harmonization to avoid
df-cols and ensure consistent factor levels across populations.
