---
title: "Compare MT content for genotyping arrays"
author: "Hansi"
date: "2025-08-07"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(proxy)
library(cluster)  
library(factoextra)
library(ggpubr)

options(warn = -1)
library(here)
library(purrr)
library(furrr)
library(progressr)
library(kableExtra)
library(plotly)
library(ggridges)
options(warn = 0)


```

## Cenotying arrays - Overview

This comparison gives an overview of the SNP counts per microarray from Will Rayners page <https://www.strand.org.uk/> as well as from the ThermoFisher (denoted here as Axiom, including the Affymetrix 6 chip).

```{r, warnings=FALSE, message=FALSE, echo=FALSE, fig.width=12, fig.height=20}


# --- 1. Load data from files with and without variants ---
illumina_data <- "data/merged_MT_data_strand2.csv"
illumina_empty <- "data/empty_microarrays_strand.txt"
axiom_summary<-read.table("../array_axiom/output/summary_table_axiom.txt")
axiom_data <- read.table("../array_axiom/output/table_axiom_full.txt")

haplogrep_annotation<-read.delim("../../reference/mtDNA/rcrs_annotated_2025_12_16.txt")
maplocus<-read.delim("../../reference/mtDNA/maplocus.txt")



# Read the merged data from files that had variants
# Handle the case where the file might not exist
if (!file.exists(illumina_data)) {
  # If the file doesn't exist, create an empty data frame
  variant_counts <- data.frame(Microarray_ID = character(), Variant_Count = numeric())
} else {
  merged_data <- read.delim(illumina_data, header = FALSE, col.names = c("Microarray_ID", "SNP_ID", "Region", "Position", "Match_Percent", "Strand", "Allele"))
  variant_counts <- merged_data %>%
    group_by(Microarray_ID) %>%
    summarise(Variant_Count = n())
}


# Read the list of empty files and create a data frame
# Handle the case where the file might not exist or be empty
if (!file.exists(illumina_empty) || file.size(illumina_empty) == 0) {
  empty_counts <- data.frame(Microarray_ID = character(), Variant_Count = numeric())
} else {
  empty_files <- readLines(illumina_empty)
  empty_microarray_ids <- gsub("_MT.txt", "", empty_files)
  empty_counts <- data.frame(Microarray_ID = empty_microarray_ids, Variant_Count = -100)
}

# --- 2. Combine all data into a single data frame ---
all_counts <- bind_rows(variant_counts, empty_counts) %>%
  arrange(desc(Variant_Count)) %>%
  # Add a status column for color coding the plot
  mutate(Status = ifelse(Variant_Count > 0, "Illumina SNPs", "Zero Variants"))

# --- 3. Generate the combined report ---

axiom_summary$V3<-"Axiom"
axiom_summary <- setNames(axiom_summary, colnames(all_counts))

all_counts_axiom <- rbind(all_counts, axiom_summary)


# Create a single, horizontal bar chart with color distinction
plot <- ggplot(all_counts, aes(x = reorder(Microarray_ID, Variant_Count), y = Variant_Count, fill = Status)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  scale_fill_manual(values = c("Illumina SNPs" = "steelblue", "Zero Variants" = "coral")) + # Use distinct colors
  labs(
    title = "MT variant counts over Illumina genotyping arrays",
    subtitle = "Zero-count microarrays are highlighted",
    x = "Microarray ID",
    y = "Number of Variants",
    fill = "Status"
  ) +
  coord_flip() + # Flip the coordinates for readability
  theme(legend.position = "bottom")

#print(plot)

# Save the plot with adjusted dimensions
ggsave("output/combined_variant_counts_report.png", plot, width = 12, height = 20, units = "in")


# Create a single, horizontal bar chart with color distinction
plot1 <- ggplot(all_counts_axiom, aes(x = reorder(Microarray_ID, Variant_Count), y = Variant_Count, fill = Status)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
   scale_fill_manual(values = c("Illumina SNPs" = "steelblue", "Zero Variants" = "coral", "Axiom"="cadetblue")) + # Use distinct colors
  labs(
    title = "MT variant counts over all genotyping arrays investigated",
    subtitle = "Zero-count microarrays are highlighted",
    x = "Microarray ID",
    y = "Number of Variants",
    fill = "Status"
  ) +
  coord_flip() + # Flip the coordinates for readability
  theme(legend.position = "bottom")

print(plot1)

# Save the plot with adjusted dimensions
ggsave("output/combined_variant_counts_axiom.png", plot1, width = 12, height = 20, units = "in")

```

## Check SNP distribution

Get an idea of how the MT-SNPs are distributed over the entire mitochondrial genome, and check if there are hotspot areas.

```{r, echo=FALSE, fig.width=12}

SNPids_df<-merged_data %>%
  group_by(SNP_ID, Position, Strand, Allele ) %>%
  select (SNP_ID, Position, Strand, Allele) %>%
  distinct(SNP_ID, Position, Strand, Allele)

write.table(SNPids_df, file="output/snpIDs.txt", row.names = FALSE, sep="\t")

merged_data$Position<-as.numeric(merged_data$Position)




combined_df <- bind_rows(
  merged_data %>% select(Microarray_ID, Position, Strand, Allele),
  axiom_data %>% mutate(Allele = paste(V13, V14, sep = "")) %>% select(Microarray_ID = V1, Position = V7, Strand=V9, Allele)
)


combined_df <- combined_df %>%
  mutate(Microarray_ID = str_remove(Microarray_ID, "-b37-strand")) %>%
  mutate(Microarray_ID = str_remove(Microarray_ID, ".b37.strand")) %>%
  mutate(Microarray_ID = str_remove(Microarray_ID, "-SourceStrand")) %>%
  mutate(Microarray_ID = str_remove(Microarray_ID, "-Source.strand")) %>%
  mutate(Microarray_ID = str_remove(Microarray_ID, "-manifest-file")) %>%
  mutate(Microarray_ID = str_remove(Microarray_ID, "Illmn.strand")) %>%
  mutate(Microarray_ID = str_remove(Microarray_ID, "SOURCE.strand")) %>%
  mutate(Microarray_ID = str_remove(Microarray_ID, ".annot")) %>%
  mutate(Microarray_ID = str_remove(Microarray_ID, "b37"))

combined_df <- combined_df %>%
  group_by(Microarray_ID) %>%
  mutate(n_snps = n_distinct(Position)) %>%
  ungroup() %>%
  mutate(id_label = paste0(Microarray_ID, " (", n_snps, ")"))

write.table(combined_df, "combined_arrays.txt", row.names = FALSE, sep="\t")

haplogrep_annotation <- haplogrep_annotation %>%
  mutate(Position = parse_number(Mutation))

haplo_unique <- haplogrep_annotation %>%
  distinct(Position, .keep_all = TRUE) %>%
  select(Position, Maplocus, Category)

merged_annotated <- combined_df %>%
  left_join(haplo_unique, by = "Position")

ggplot(merged_annotated, aes(x=Maplocus)) + geom_bar(aes(fill=Category)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(legend.position = "bottom") +ggtitle("All variants over all arrays annotated with locus")

maplocus_summary <- merged_annotated %>%
     group_by(Maplocus, Category) %>%
     summarise(
         amount = n()
     )

maplocus_merged<-merge(maplocus_summary, maplocus, by.x="Maplocus", by.y="MapLocus")
maplocus_merged$sizeadjusted<-maplocus_merged$amount/maplocus_merged$length

ggplot(maplocus_merged, aes(x=Maplocus, y=sizeadjusted)) + geom_point(aes(color=Category)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(legend.position = "bottom") +ggtitle("All variants over all arrays annotated with locus", subtitle="Gene-size adjusted")

maplocus_merged_cat <- maplocus_merged %>%
     group_by(Category) %>%
     summarise(
         Catsum = sum(amount), 
        Catlength=sum(length)   
     )

ggplot(maplocus_merged_cat, aes(x=Category, y=Catsum/Catlength)) + geom_point(aes(color=Category)) + stat_compare_means() + ylim(0,4)  +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(legend.position = "bottom")+ggtitle("All variants over all arrays grouped by majo locus-category", subtitle="Gene-size adjusted")


```

## Clustering Analysis of Arrays

Preparing a matrix with all microarrays and all possible Variants observed over all arrays we calculate the Jaccard distance

```{r, echo=FALSE, fig.width=18, fig.height=12}
counts_table <- table(combined_df$id_label, combined_df$Position)
binary_matrix <- as.data.frame.matrix(counts_table > 0) + 0


jac_dist <- proxy::dist(as.matrix(binary_matrix), method = "Jaccard")

# Hierarchical Clustering
hc <- hclust(jac_dist, method = "ward.D2")

smc_dist <- proxy::dist(as.matrix(binary_matrix), method = "Simple Matching")
hc_smc <- hclust(smc_dist, method = "ward.D2")


fviz_nbclust(as.matrix(binary_matrix), hcut, method = "silhouette", hc_method = "ward.D2")

# 5. Visualization
# This creates a dendrogram to see design similarity

glasbey_colors <- c(
  "#0000FF", "#FF0000", "#00FF00", "#000033", "#FF00B6", 
  "#005300", "#FFD300", "#009FFF", "#9A4D42", "#aac000"
)

p<-fviz_dend(hc, k =10, k_colors=glasbey_colors, cex = 0.6, type="rectangle", main = "Cluster Dendrogram of Microarray Designs",xlab = "Microarray ID",  ylab = "Jaccard Distance", labels_track_height = 2) 

p2<-fviz_dend(hc_smc, k =10, k_colors=glasbey_colors, cex = 0.6, type="rectangle", main = "Cluster Dendrogram of Microarray Designs",xlab = "Microarray ID",  ylab = "Jaccard Distance", labels_track_height = 2) 

plot(p)


ggsave("output/dendrogram_jaccard.pdf", p, width = 18, height = 12)
ggsave("output/dendrogram_simple_matching.pdf", p2, width = 18, height = 12)


##### FILTER for identical groups:

# 1. Convert to data frame to use dplyr's distinct()

binary_df <- binary_matrix
binary_df$ArrayType <- rownames(binary_matrix)


# 2. Keep only one representative for each unique pattern of 1s and 0s
unique_binary_df <- binary_df %>%
  distinct(across(-ArrayType), .keep_all = TRUE)

# 3. Set row names back to the ArrayType and remove the helper column
rownames(unique_binary_df) <- unique_binary_df$ArrayType
unique_matrix_final <- as.matrix(unique_binary_df %>% select(-ArrayType))

cat("Original microarrays:", nrow(binary_matrix), "\n")
cat("Unique fingerprints:", nrow(unique_matrix_final), "\n")

# Calculate Jaccard distance for the unique set
jac_dist_unique <- proxy::dist(unique_matrix_final, method = "Jaccard")

# Re-run the silhouette analysis
# This plot will show you a "peak" at the ideal number of clusters
fviz_nbclust(unique_matrix_final, hcut, 
             method = "silhouette", 
             hc_method = "ward.D2") +
  labs(title = "Optimal Number of Clusters (Unique Fingerprints Only)")


# Run hierarchical clustering
hc_unique <- hclust(jac_dist_unique, method = "ward.D2")

p_unique<-fviz_dend(hc_unique, k = 10, 
          cex = 0.5, # Size of the labels
          k_colors = "jco", 
          rect = TRUE, # Add rectangles around clusters
          rect_fill = TRUE,
          main = "Hierarchical Clustering (k=10)")

plot(p_unique)



# Cut the tree into your chosen number of clusters (e.g., 10)
k_choice <- 10 
unique_clusters <- cutree(hc_unique, k = k_choice)

# Create a mapping of the unique clusters
unique_cluster_df <- data.frame(
  ArrayType = names(unique_clusters),
  ClusterID = as.factor(unique_clusters)
)

# These are your 10 most distinct representatives
representative_10 <- unique_cluster_df %>%
  group_by(ClusterID) %>%
  slice(1) %>%
  pull(ArrayType)

print(representative_10)


#### Select 10 representative arrays:

# Cut the tree into 10 clusters
clusters <- cutree(hc, k = 10)

# Create a clean data frame of the results
cluster_assignment <- data.frame(
  Microarray = names(clusters),
  ClusterID = as.factor(clusters)
)

# See the names in Cluster 1, for example:
subset(cluster_assignment, ClusterID == 1)

# Or get a list of names grouped by cluster
split(cluster_assignment$Microarray, cluster_assignment$ClusterID)

representative_samples <- cluster_assignment %>%
  group_by(ClusterID) %>%
  slice(1) %>% # This picks the first microarray appearing in each cluster
  pull(Microarray)

print(representative_samples)



final_representatives <- unique_cluster_df %>%
  mutate(
    snp_num = as.numeric(str_extract(ArrayType, "(?<=\\()\\d+(?=\\))")),
  ) %>%
  filter(snp_num >= 10) %>% 
  group_by(ClusterID) %>%
  # Calculate the mean and identify which row is closest to it
  mutate(
    cluster_mean = mean(snp_num, na.rm = TRUE),
    is_closest_to_mean = abs(snp_num - cluster_mean) == min(abs(snp_num - cluster_mean))
  ) %>%
  # Filter for min, max, OR the one closest to the mean
  filter(snp_num == min(snp_num) | 
         snp_num == max(snp_num) | 
         is_closest_to_mean) %>%
  # Label them accordingly
  mutate(extreme_label = case_when(
    snp_num == max(snp_num) ~ "Highest",
    snp_num == min(snp_num) ~ "Lowest",
    is_closest_to_mean      ~ "Mean-ish"
  )) %>%
  # Handle ties or overlaps (e.g., if the mean is also the max)
  distinct(ClusterID, extreme_label, .keep_all = TRUE) %>%
  ungroup()

# Quick check: did we get the right numbers?
print(select(final_representatives, ArrayType, ClusterID, snp_num, extreme_label))

write.table(final_representatives, file="final_representative_arrays.txt", sep=",", row.names = FALSE)
write.table(unique_cluster_df, file="all_unique_arrays.txt", sep=",", row.names = FALSE)
```

## Check RefAlt files

Use SNP-IDs prepared previously to run over all .RefAlt files downloaded from here: <https://www.chg.ox.ac.uk/~wrayner/strand/RefAlt/> and unzipped in own folder with mergeSnpIDs.py - this was the first version in assessing the SNPs present.

```{r, warnings=FALSE, message=FALSE, echo=FALSE }
refAlt_df <- read.delim("data/matched_snps.RefAlt.txt")


variant_counts <- refAlt_df %>%
  group_by(Filename) %>%
  summarise(num_variants = n()) %>%
  arrange(desc(num_variants))

p<-ggplot(variant_counts, aes(x = num_variants, y = reorder(Filename, num_variants))) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Number of Variants per Filename in RefAlt files",
    x = "Variant Count",
    y = "Filename"
  ) + theme_minimal()

length(unique(refAlt_df$Filename))

snp_refAlt_df<-merge(refAlt_df, SNPids_df, by.x="SNP_ID", by.y="SNP_ID", all.x=TRUE)

snp_refAlt_pos_df<-merge(refAlt_df, SNPids_df[SNPids_df$Strand=="+",], by.x="SNP_ID", by.y="SNP_ID", all.x=TRUE)
snp_refAlt_pos_df$Position<-as.numeric(snp_refAlt_pos_df$Position)

snp_refALT_pos_filt_df <- snp_refAlt_pos_df %>%
  select(SNP_ID, Filename, Position) %>%
  distinct(SNP_ID, Filename, Position)

snp_refAlt_neg_df<-merge(refAlt_df, SNPids_df[SNPids_df$Strand=="-",], by.x="SNP_ID", by.y="SNP_ID", all.x=TRUE)




```

# Compare major arrays for haplogroup metrics

```{r, echo=FALSE, warnings=FALSE, message=FALSE, fig.width=9, fig.height=11, out.width="100%"}

source(here("scripts/haplogroup_processing_dataframe.R"))
compute_path <- here("scripts", "haplogroup_processing_dataframe.R")

phylotreeVersion <- "17_FU1"



.options = furrr_options(
  globals = c("process_haplogroup_data", "hsd_path_abs"),
  packages = c("dplyr", "stringr", "networkD3", "htmlwidgets")
)

#args(process_haplogroup_data)
#readLines("../../scripts/haplogroup_processing.R", n = 100)

##test-scenarios######################################################
#hsd_path_abs <- normalizePath("data/test10AFR.hsd", mustWork = TRUE)
hsd_path_abs <- normalizePath("data/test193AFR.hsd", mustWork = TRUE)
hsd_data <- read.delim(hsd_path_abs)
hsd_data[[3]] <- ""
hsd_data <- hsd_data[, colSums(is.na(hsd_data)) == 0]

pop="AFR_193"


#csv_path_abs <- normalizePath("../../reference/output/ref_AFR.csv", mustWork = TRUE)
#csv_path_abs <- normalizePath("../../reference/output/ref_EUR.csv", mustWork = TRUE)
#csv_path_abs <- normalizePath("../../reference/output/ref_AMR.csv", mustWork = TRUE)
#csv_path_abs <- normalizePath("../../reference/output/ref_EAS.csv", mustWork = TRUE)
#csv_path_abs <- normalizePath("../../reference/output/ref_SAS.csv", mustWork = TRUE)
#csv_path_abs <- normalizePath("../../reference/output/ref_ALL.csv", mustWork = TRUE)
#csv_path_abs <- normalizePath("../../reference/output/ref_Oth.csv", mustWork = TRUE)
#csv_data <- read.delim(csv_path_abs, sep=",")
#hsd_data <- csv_data[,1:4]
#hsd_data$SampleID<-paste(hsd_data$SampleID,"-",hsd_data$Haplogroup, sep="")
#hsd_data[[3]] <- ""
#hsd_data <- hsd_data[, colSums(is.na(hsd_data)) == 0]
#hsd_data <- hsd_data[hsd_data$SampleID != "mt-MRCA", ]


#plan(sequential)
#closeAllConnections()

groups <- split(combined_df, combined_df$Microarray_ID)

# Filter based on clustering analysis
rep_names_full <- trimws(str_remove(final_representatives$ArrayType, "\\s*\\(.*"))
test_groups <- groups[names(groups) %in% rep_names_full]

#Filter based on unique arrays - basically run over all
rep_names_unique <- trimws(str_remove(unique_cluster_df$ArrayType,  "\\s*\\(.*"))
test_groups <- groups[names(groups) %in% rep_names_unique]


plan(multisession, workers = parallel::detectCores() - 2)

handlers("progress")

with_progress({
  haplogroupsDiff_all <- future_map_dfr(
    test_groups,
    function(df_group) {
      current_id <- unique(df_group$Microarray_ID)
      process_haplogroup_data(
        array_type = current_id,
        array_df = df_group,
        hsd_data = hsd_data,
        pop=pop,
        output_base_name = paste0("out_", current_id)
      )
    },
    .options = furrr_options(
      globals = c(
        "process_haplogroup_data",
        "hsd_data",
        "pop",
        "phylotreeVersion",
        "get_Supergroups",
        ".derive_MacroPhylo"
      ),
      packages = c("dplyr", "stringr", "networkD3", "htmlwidgets", "here")
    ),
    .progress = TRUE
  )
})

p <- ggplot(haplogroupsDiff_all, aes(x = distance, color = ArrayType, group = ArrayType)) +
  stat_ecdf(geom = "line") +
  theme_minimal()

ggplotly(p)


ggplot(haplogroupsDiff_all, aes(x = distance, y = ArrayType, fill = ..x..)) +
  geom_density_ridges_gradient(scale = 3, rel_min_height = 0.01) +
  scale_fill_viridis_c(name = "Distance", option = "C") +
  labs(title = "Distributions of Distances across Genotyping Arrays") +
  theme_ridges() +
  theme(axis.text.y = element_text(size = 6)) # Shrink text to fit 150 labels


ggplot(haplogroupsDiff_all, aes(x = distance, group = ArrayType)) +
  # Draw all 150 lines in very light gray
  stat_ecdf(geom = "line", color = "gray90", size = 0.2) + 
  # Highlight the global average or a specific group of interest in bold
  stat_ecdf(data = haplogroupsDiff_all, aes(x = distance, group = 1), 
            geom = "line", color = "deepskyblue1", linewidth = 1) +
  theme_minimal() +
  labs(title = "Individual Haplogroup Distance vs. Global Average (Blue)")


# 1. Calculate percentiles for each ArrayType
# We'll look at every 10th percentile plus the 95th and 99th for outlier detection
heatmap_data <- haplogroupsDiff_all %>%
  group_by(ArrayType) %>%
  reframe(
    percentile = c(seq(0.1, 0.9, by = 0.1), 0.95, 0.99),
    dist_value = quantile(distance, probs = percentile, na.rm = TRUE)
  )

# 2. Create the Heatmap
ggplot(heatmap_data, aes(x = factor(percentile), 
                         y = reorder(ArrayType, dist_value, FUN = median), 
                         fill = dist_value)) +
  geom_tile() +
  # Using the 'magma' palette which is great for seeing density/intensity
  scale_fill_viridis_c(option = "magma", name = "Observed\nDistance") +
  labs(
    title = "Haplogroup Distance Fingerprints",
    subtitle = "Groups sorted by median distance",
    x = "Percentile (0.5 = Median)",
    y = "Array Type (n=150)"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    # With 150 rows, the Y-axis text must be very small
    axis.text.y = element_text(size = 5), 
    axis.text.x = element_text(angle = 45, hjust = 1)
  )


# Calculate means and grab top 20
top_20 <- haplogroupsDiff_all %>%
  group_by(ArrayType) %>%
  summarize(mean_dist = mean(distance, na.rm = TRUE)) %>%
  slice_max(mean_dist, n = 20)

# Plot with reorder() to break alphabetical sorting
ggplot(top_20, aes(x = mean_dist, y = reorder(ArrayType, mean_dist))) +
  geom_col(fill = "firebrick") +
  labs(
    title = "Worst 20 Array Types by Highest Mean Distance",
    x = "Mean Observed Distance (lower=better)",
    y = "Array Type"
  ) +
  theme_minimal()


# Calculate means and grab bottom 20
bottom_20 <- haplogroupsDiff_all %>%
  group_by(ArrayType) %>%
  summarize(mean_dist = mean(distance, na.rm = TRUE)) %>%
  slice_min(mean_dist, n = 20)

# Plot with reorder() - negative mean_dist sorts it from lowest to highest
ggplot(bottom_20, aes(x = mean_dist, y = reorder(ArrayType, -mean_dist))) +
  geom_col(fill = "steelblue") +
  labs(
    title = "Best 20 Array Types by Lowest Mean Distance",
    x = "Mean Observed Distance (lower = better)",
    y = "Array Type"
  ) +
  theme_minimal()


#######################################

# 1. Rank your ArrayTypes by performance (e.g., Median distance)
ranking <- haplogroupsDiff_all %>%
  group_by(ArrayType) %>%
  summarize(med = median(distance, na.rm = TRUE)) %>%
  arrange(med)

top_5 <- head(ranking$ArrayType, 5)
bottom_5 <- tail(ranking$ArrayType, 5)
keep_in_legend <- c(top_5, bottom_5)

# 2. Add a 'ShowInLegend' flag and 'CustomColor' to your data
haplogroupsDiff_all <- haplogroupsDiff_all %>%
  mutate(
    Group = ifelse(ArrayType %in% keep_in_legend, as.character(ArrayType), "Other"),
    InLegend = ArrayType %in% keep_in_legend
  )

# 3. Build the Plot
p <- ggplot(haplogroupsDiff_all, aes(x = distance, color = Group, group = ArrayType)) +
  # Use 'text' aesthetic to customize the mouse-over tooltip
  stat_ecdf(geom = "line", aes(text = paste("Array:", ArrayType)), lwd = 0.8) +
  scale_color_manual(values = c(
    setNames(scales::hue_pal()(length(keep_in_legend)), keep_in_legend), 
    "Other" = "grey85"
  )) +
  scale_x_continuous(breaks = 0:25) +
  labs(title = "Interactive Haplogroup Distribution", x = "Distance", y = "Cumulative Prop") +
  theme_minimal() + theme(legend.position = "bottom")

# 4. Convert to Plotly and fix the legend
# We hide the legend for 'Other' items to keep it clean
fig <- ggplotly(p, tooltip = "text") %>%
  style(showlegend = FALSE, traces = which(unique(haplogroupsDiff_all$Group) == "Other"))

fig

######################################
### LIMIT THE ARRAYS to 10 Clusters 

# Join back to your main distance data

final_representatives <- final_representatives %>%
  mutate(ArrayType = clean_text <- sub("\\s*\\(.*", "", ArrayType)) 


rep_20_plot_data <- haplogroupsDiff_all %>%
  filter(ArrayType %in% final_representatives$ArrayType) %>%
  left_join(final_representatives, by = "ArrayType")

ggplot(rep_20_plot_data, aes(x = distance, color = extreme_label, group = ArrayType)) +
  stat_ecdf(geom = "line", size = 1.1) +
  facet_wrap(~ClusterID, ncol = 5, labeller = label_both) + 
  scale_color_manual(values = c("Highest" = "#D55E00", "Lowest" = "#0072B2", "Mean-ish"="#009E73")) +
  labs(
    title = "Comparison of Cluster Extremes",
    subtitle = "Comparing arrays with the largest vs. smallest bracketed ID within each cluster",
    x = "Observed Haplogroup Distance",
    y = "Cumulative Proportion",
    color = "Intra-Cluster Rank"
  ) +
  theme_minimal() +
  theme(strip.background = element_rect(fill = "gray95", color = NA),
        legend.position = "bottom")

#######################################



max_dist <- max(haplogroupsDiff_all$distance)
dist_range <- 0:max_dist
num_points <- length(dist_range)

auc_results <- haplogroupsDiff_all %>%
  group_by(ArrayType) %>%
  summarise(
    AUC_Normalized = sum(ecdf(distance)(dist_range)) / num_points,
    Mean_Distance = mean(distance, na.rm = TRUE),
    n = n()
  ) %>%
  mutate(Tier = case_when(
    AUC_Normalized >= 0.95 ~ "🥇 Gold (Excellent)",
    AUC_Normalized >= 0.85 ~ "🥈 Silver (Good)",
    AUC_Normalized >= 0.75 ~ "🥉 Bronze (Fair)",
    TRUE                   ~ "⚠️ Low Performance"
  )) %>%
  arrange(desc(AUC_Normalized))

# Update names for the Sankey section
best_array_name  <- auc_results$ArrayType[1]
worst_array_name <- auc_results$ArrayType[nrow(auc_results)]


kable(auc_results, 
      col.names = c("Microarray", "Normalized AUC", "Mean Distance", "Sample Size", "Performance Tier"),
      digits = 3,
      caption = "Microarray Performance Ranking with Tiers")

ggplot(auc_results, aes(x = reorder(ArrayType, AUC_Normalized), y = AUC_Normalized, fill = Tier)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("🥇 Gold (Excellent)" = "#FFD700", 
                               "🥈 Silver (Good)" = "#C0C0C0", 
                               "🥉 Bronze (Fair)" = "#CD7F32", 
                               "⚠️ Low Performance" = "#E57373")) +
  labs(title = "Microarray Accuracy Scores (AUC Normalized)",
       x = "Microarray Type",
       y = "Accuracy Score (0-1)",
       fill = "Tier") +
  theme_minimal()

```


